Fourier-domain modulations and delays of gravitational-wave signals

### 2018-06-28 docs/paper_method.tex
First submitted version of the paper.

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{color}
\usepackage{enumerate}
\usepackage{ulem}
\usepackage{tabularx}
\usepackage{braket}

@@

\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand\ud{{\mathrm{d}}}
\newcommand\uD{{\mathrm{D}}}
\newcommand\calO{{\mathcal{O}}}
\newcommand\calM{{\mathcal{M}}}
\newcommand\calF{{\mathcal{F}}}
\newcommand\calT{{\mathcal{T}}}
\newcommand\calD{{\mathcal{D}}}
\newcommand\calA{{\mathcal{A}}}
\newcommand\bfx{\mathbf{x}}
\newcommand{\ov}[1]{\overline{#1}}
\newcommand{\ph}[1]{\phantom{#1}}
\newcommand{\cte}{\mathrm{cte}}
\newcommand{\nn}{\nonumber}
%\newcommand{\hatk}{\hat{k}}
\newcommand{\hatk}{k}
\newcommand{\Hz}{\,\mathrm{Hz}}
\newcommand{\yr}{\,\mathrm{yr}}
\newcommand{\sinc}{\,\mathrm{sinc}}
\newcommand{\Msol}{M_{\odot}}
\newcommand{\Mchirp}{M_{c}}
\newcommand{\tf}{t_{f}}
\newcommand{\Tf}{T_{f}}
\newcommand{\tfd}{t_{f}^{d}}
\newcommand{\tfSPA}{t_{f}^{\rm SPA}}
\newcommand{\TfSPA}{T_{f}^{\rm SPA}}

@@

We present a Fourier-domain based approach modulations and delays of gravitational wave signals, a problem which arises in two different contexts. For space-based detectors like LISA, the orbital motion of the detector introduces a time-dependency in the response of the detector, consisting of both a modulation and a varying delay. In the context of signals from precessing spinning binary systems, a useful tool for building models of the waveform consists in representing the signal as a time-dependent rotation of a quasi-non-precessing waveform. In both cases, being able to compute transfer functions for these effects directly in the Fourier domain may enable performance improvements for data analysis applications by using fast frequency-domain waveforms. Our results generalize previous approaches based on the stationary phase approximation for inspiral signals, extending them by including delays and computing corrections beyond the leading order, while being applicable to the broader class of signals of inspiral-merger-ringdown waveforms. In the LISA case, we find that a leading-order treatment is accurate for high-mass and low-mass signals that are chirping enough, with reconstruction errors that can be reduced by consistently including the corrections we derived. By contrast, low-mass binary black holes, if far away from merger and slowly-chirping, cannot be handled by our perturbative formalism and we develop another approach for these systems. In the case of precessing binaries, we perform a limited exploration of the merger-ringdown range based on a toy model for the precession. We find that, depending on the post-merger frame velocity, deviations from leading order can become qualitatively large to the point of challenging our perturbative formalism, while affecting mainly subdominant mode contributions and giving rise to a limited unfaithfulness in the full waveform. Including higher-order corrections consistently reduces the unfaithfulness, and we further develop an alternative approach to accurately represent post-merger features.

@@

With the unprecedented recent gravitational-wave detections of coalescencing binary black holes and binary neutron stars, announced by the LIGO-Virgo collaboration~\cite{LIGO-theevent-2016, LIGO-christmasevent-2016, LIGO-O1BBH-2016, LIGO-BNSevent-2017}, gravitational-wave astronomy has entered its observational era. As LIGO prepares for even more sensitive observation runs, and with the recent expansion of the ground-based detectors network with Virgo~\cite{Virgo} (and eventually also KAGRA~\cite{KAGRA} and LIGO-India~\cite{INDIGO}), observations of such compact object coalescences are expected at ever-increasing rate.

Moreover, the European Space Agency has recently selected the Laser Interferometer Space Antenna (LISA)\cite{LISA17} to realize the ``Gravitational Universe'' science theme\cite{elisa13} as the 3rd large space mission of its Cosmic Vision program, with a tentative launch around 2034. LISA will be able to detect and characterize, among several important gravitational wave source targets, comparable-mass binary black hole coalescences from cosmological distances over a wide range of masses. These will range from high-redshift observations of supermassive black hole binaries with $M\sim 10^{7} \Msol$ down to the observation of LIGO-type sources with $M\sim 10^{2} \Msol$~\cite{Sesana16}.

@@

%Our challenge will be to accurately complete the representation of instrumental signals including precession effects and the LISA instrumental response and precession effects without sacrificing the computational efficiency enabled by these fast waveform templates. \SM{[rewrite that]}

A complete representation of spin effects in fast IMR templates still remains a frontier of gravitational wave signal modelling. In presence of misaligned spins, the system will endure precession of the orbital plane as it evolves, leading to modulations of the signal as seen by the observer~\cite{Apostolatos+94, Kidder95}. The presence of six degrees of freedom for the spins increases the dimensionality of the problem.

@@

%Prospective parameter recovery studies for LISA observations of comparable-mass binary black hole coalescences aim at characterizing the scientific performance of the future experiment. Although LISA~\cite{LISA17} is still at least a decade from launch, the ESA L3 mission is rapidly progressing into its early stages where mission design choices must be informed by science impact analysis. However, the recent progresses in waveform modelling, most notably the availability of fast and accurate IMR waveforms, have not yet made their way to the field of LISA data analysis studies. In the past, many such studies used post-Newtonian, inspiral-only models, and were limited to the cheaper Fisher-matrix formalism instead of a full Bayesian analysis (among numerous works, see e.g. \cite{Cutler97,LISAPE09}). Refs.~\cite{McWilliams+09,McWilliams+10,McWilliams+11} explored the effects of including the merger and ringdown, keeping the mass parameters fixed. However, the Fisher matrix analysis, although considerably cheaper to perform, is valid only in the limit of a high signal to noise ratio (SNR). Several implementations of Bayesian analyses were developed~\cite{CP06,Wickham+06,Babak+08,GP09,Feroz+09,Petiteau+10,PC13} (see also~\cite{MLDC4} and references therein), with Ref.~\cite{PC15} proposing a direct comparison of Fisher and Bayes analyses, but they were limited to inspiral-only waveforms. One exception was Ref.~\cite{Babak+08} where IMR signals were used in a Bayesian study, but there the masses of the black holes were kept fixed. In the recent Ref.~\cite{Klein+15}, the limitation to inspiral-only precessing post-Newtonian waveforms was compensated by an SNR reweighting procedure. Thus, due mainly to their high computational cost, full-fledged Bayesian parameter estimation simulations for IMR signals are still missing. \jgb{[Maybe there is too much detail in this paragraph?]}

%Prospective parameter recovery studies for LISA observations of comparable-mass binary black hole coalescences aim at characterizing the scientific performance of the future experiment. Although LISA~\cite{LISA17} is still at least a decade from launch, the ESA L3 mission is rapidly progressing into its early stages where mission design choice must be informed by science impact analysis. However, the recent progresses in waveform modelling, most notably the availability of cheap IMR waveforms, have not yet made their way to the field of LISA data analysis studies. In the past, such studies used mainly post-Newtonian, inspiral-only models, and were limited to the Fisher-matrix formalism instead of full Bayesian analysis (among numerous works, see e.g. \cite{Cutler97,LISAPE09}). Refs.~\cite{McWilliams+09,McWilliams+10,McWilliams+11} explored the effects of including the merger and ringdown, keeping the mass parameters fixed. However, the Fisher matrix analysis, although considerably cheaper to perform, is valid only in the limit of a high signal to noise ratio (SNR) and has known shortcomings~\cite{Vallisneri08}. Several implementations of Bayesian analyses were developed~\cite{CP06,Wickham+06,Babak+08,GP09,Feroz+09,Petiteau+10,PC13} (see also~\cite{MLDC4} and references therein), with Ref.~\cite{PC15} proposing a direct comparison of Fisher and Bayes analyses, but they were limited to inspiral-only waveforms. One exception was Ref.~\cite{Babak+08} where IMR signals were used in a Bayesian study, but there the masses of the black holes were kept fixed. In the recent Ref.~\cite{Klein+15}, the limitation to inspiral-only precessing post-Newtonian waveforms was compensated by an SNR reweighting procedure. Thus, due mainly to their high computational cost, full-fledged Bayesian parameter estimation simulations for IMR signals are still missing. \jgb{[Maybe there is too much detail in this paragraph?]}

Whereas LIGO and Virgo are typically sensitive to chirping binaries for a minute or less, LISA signals may accumulating over months or years. The response of a LISA-type instrument is thus time-dependent~\cite{Cutler97}. The motion and change of orientation of the detector constellation along its orbit lead to significant time variability in the form of a modulation and a varying delay. These effects then convey information about the localization of the gravitational-wave source in the sky. Direct time-domain implementation of the detector response is straightforward~\cite{Vallisneri04, Petiteau+08, CR02, RCP04}, but at a prohibitive computational cost for parameter-estimation analyses. To leverage the performance of state-of-the-art Fourier-domain IMR templates~\cite{BTB16,Khan+15}, we must efficiently process the signals through the time-dependent response of the dectector while staying in the Fourier domain.

@@

In the case of precessing binaries, applying the SPA directly to the modulated signal is also prone to pathologies. In Refs.~\cite{KCY13,KCY14}, a formalism (called shifted uniform asymptotics or SUA) was introduced to go beyond the SPA and compute more accurately the modulation in the Fourier domain; however, this formalism still relies on the SPA for the underlying precessing-frame signal, and is as such limited to inspiraling signals. The simplified treatment of the precession response in the phenomenological waveforms of Ref.~\cite{Hannam+13} takes another approach, treating the precession modulation in the frequency domain by directly associating the Fourier frequency with the post-Newtonain orbital frequency. As we will explain, this corresponds to the zeroth-order approximation of the SUA.

In the case of the LISA response, the SPA provides a natural map from time-domain (for the orbit) to Fourier-domain (for the signal)~\cite{Cutler97}, and was used by many previous studies with inspiral waveforms. Ref.~\cite{Klein+15} included the orbital motion of the detector in the SUA treatment of precessing inspiral signals, within a low-frequency approximation of the constellation response. Consistently extending these previous approaches to the merger-ringdown part of the signals, while including the delays in the full LISA response at all frequencies, is part of the objectives of this paper.

@@

The plan of the paper is as follows. In Sec.~\ref{sec:motivation}, we provide a general presentation of the problem of Fourier-domain modulation and introduce the relevant timescales for both the response of LISA-type detectors and the modulation of precessing signals. In Sec.~\ref{sec:formalism}, we present our general formalism, give its leading order approximation as well as higher-order corrections, introduce new timescales based on the Fourier-domain signal, refine the previous results for both the quadratic-in-phase corrections and the treatment of the delays, and make the connection to previous approaches of the problem. We then apply our formalism to the response of the LISA detector in Sec.~\ref{sec:LISA}, and to the case of signals from precessing binaries in Sec~\ref{sec:precession}. We discuss and summarize our results in Sec.~\ref{sec:discussion}.

@@

%\subsection{Basic considerations}
%\label{subsec:basic}

@@

As is well known, GW signals decomposed in spin-weighted spherical harmonic components are smoothly varying functions of frequency in amplitude/phase form. An important consequence is that these signal components can be accurately represented in terms of a relatively coarsely sampled frequency grid. This property will also extend to the transfer functions, allowing us to keep a compact representation of the full signals. (See Appendix~\ref{app:notation} for details of our notation and conventions for Fourier transforms and spherical harmonics.)

@@

As a preliminary step, we recall here an approximation widely used for similar purposes in treating gravitational waves signals emitted by inspiraling binaries, the Stationary Phase Approximation (thereafter SPA, and also sometimes called the steepest-descent method). It applies in general to chirping signals, and we refer the reader to~\cite{FC93, CF94} for details. One first writes the time-domain signal in amplitude and phase form as $h(t) = a(t) e^{-2i\varphi(t)}$, where for gravitational wave signals $\varphi$ will correspond to the orbital phase, with the orbital frequency being $\omega = \dot{\varphi}$. In order to keep close to the notation used in the gravitational wave literature, we introduced a factor of 2 in the phase of the wave, which is appropriate for the dominant 22 harmonic of the signal. The approximation then applies to signals verifying the conditions\footnote{Note that the third condition is not always written explicitly, in particular not in Refs.~\cite{FC93, CF94}. It becomes important if we generalize $a$ to be an envelope function incorporating a modulation.}
\be\label{eq:conditionsSPA}
	\left| \frac{\dot{a}/a}{\omega} \right| \ll 1\,, \quad \left|\frac{\dot{\omega}}{\omega^{2}} \right| \ll 1\,, \quad \left| \frac{(\dot{a}/a)^{2}}{\dot{\omega}} \right| \ll 1 \,.
\ee
Since the integral~\eqref{eq:defFT} defining the Fourier transform is rapidly oscillatory unless the term $2\pi f t$ cancels the evolution of $-2\varphi(t)$, its support is well centered around the point of stationary phase. This determines the time-to-frequency correspondence in the stationary phase approximation, and leads to the definition of this time as an implicit function of frequency by the relation
\be\label{eq:deftfSPA}
	\omega(\tfSPA) = \pi  f \,.
\ee
For a chirping signal of increasing phase, $\omega>0$ and $\dot{\omega}>0$, there is a unique point of stationary phase, located in the positive frequency range $f>0$. Using the conditions above, one can formally expand the signal around $\tfSPA$ to quadratic order in time, according to
\begin{align}
	\tilde{h}_{\rm SPA} (f) &\simeq a(\tfSPA) \exp\left[2i\pi f \tfSPA-2i\varphi(\tfSPA) \right] \nn\\
	& \qquad \cdot \int \ud t \, e^{-i \dot{\omega} (\tfSPA) (t-\tfSPA)^{2}} \,.
\end{align}
Note that to be able to treat the amplitude as a constant in the integral above, we used the third condition in~\eqref{eq:conditionSPA}. The resulting complex Gaussian integral yields\footnote{The expressions given here are valid for the dominant mode $h_{22}$ of the waveform. They can be generalized to other modes $h_{\ell m}$, $m\neq 0$ with phase $e^{-im\varphi}$ as follows: $\Psi$ acquires a factor $m/2$, $A$ a factor $\sqrt{2/m}$, and $v$ (see~\eqref{eq:SPAN}) has to be replaced by $(2/m)^{1/3} v$.}
\begin{subequations}
\begin{align}
	\tilde{h}_{\rm SPA}(f) &= A_{\rm SPA}(f) e^{-i\Psi_{\rm SPA}(f)} \,, \\
	A_{\rm SPA}(f) &= a(\tfSPA) \sqrt{\frac{\pi}{\dot{\omega}(\tfSPA)}} \,, \label{eq:ASPA} \\
	\Psi_{\rm SPA}(f) &= 2\varphi(\tfSPA) - 2\pi f \tfSPA + \frac{\pi}{4} \,. \label{eq:PsiSPA}
\end{align}
\end{subequations}
Ref.~\cite{Droz+99} evaluated the first correction to this approximation, within the context of post-Newtonian signals, and found that it can be considered as a term of the fifth post-Newtonian order, beyond the accuracy level of our current best models~\cite{BlanchetLiving}.

To understand the separation of timescales in the problem, it will be useful to have at hand the leading-order scaling laws for an inspiral (labeled the Newtonian order in the PN language). Although inaccurate for the purpose of waveform modelling, these leading-order estimates will give useful orders of magnitude of the relevant timescales in the inspiral. For a binary with masses $m_{1}, m_{2}$, we define the total mass $M=m_{1}+m_{2}$ and the symmetric mass ratio $\nu = m_{1}m_{2}/M^{2}$. Introducing a time of coalescence $t_{c}$, the relations between the orbital frequency and phase and the time to coalescence $t_{c} - t$ are then given by
\begin{subequations}\label{eq:omegaphiN}
\begin{align}
	\omega(t) &= \left[ \frac{256\nu}{5c^{5}} (GM)^{5/3} (t_{c}-t) \right]^{-3/8} \,, \\
	\varphi(t) &= -\left[ \frac{c^{3}}{5 G M \nu^{3/5}} (t_{c}-t) \right]^{5/8} \,.
\end{align}
\end{subequations}
As for the leading-order time-domain amplitude of the $22$ mode, we have~\cite{BlanchetLiving}
\be\label{eq:a22N}
	a_{22}^{\rm N} (t) = \frac{2 G M \nu v^{2}}{D c^{2}} \sqrt{\frac{16 \pi}{5}} \,,
\ee
where we set $v = (G M\omega/c^{3})^{1/3}$ and where $D$ is the luminosity distance to the observer. For this Newtonian inspiral, applying the SPA gives for the $22$ mode $\tilde{h}(f) = A_{\rm N}(f)e^{-i\Psi_{\rm N}(f)}$ with
\begin{subequations}\label{eq:SPAN}
\begin{align}
	A_{N}(f) &= \frac{G^{2}M^{2} \pi}{Dc^{5}} \sqrt{\frac{2\nu}{3}} v^{-7/2}\,, \label{eq:ASPAN}\\
	\Psi_{\rm SPA}^{\rm N}(f) &= \phi_{0} - 2\pi f t_{0} - \frac{3}{128\nu v^{5}} \,, \label{eq:PsiSPAN} 
\end{align}
\end{subequations}
where $v=(G M \pi f/c^{3})^{1/3}$ according to the SPA correspondence~\eqref{eq:deftfSPA}, and where $t_{0}, \phi_{0}$ are constants. It is also customary to rewrite the above relations in terms of the chirp mass $\Mchirp \equiv M\nu^{3/5}$, which is the only mass combination characterizing the signal at the leading PN order.

In this Newtonian, low-frequency limit one can check that each of the combinations \eqref{eq:conditionsSPA} indeed vanish at $\calO{(v^5)}$. but as the system approaches merger, these condition are no-longer satisfied. Near and after the merger, the SPA treatment is not applicable.
%, as can be seen by the breakdown of the relation~\eqref{eq:deftfSPA} as $\omega$ asymptotes the quasi-normal mode frequency in the ringdown. Because of this limitation, we aim at developing a formalism that would not rely on this approximation.
Nonetheless, the SPA has been a useful workhorse in many gravitational analyses involving frequency domain transfer functions of the form~\eqref{eq:deftransfer}. The usual approach is simply to replace any time dependencies appearing the transfer function using \eqref{eq:deftfSPA}. We will reference the SPA treatment, as a familiar touchstone, as we develop a more general formalism.

@@

For the signal, then, we have
\be
	\tilde{h}(f-f') \simeq A(f) \exp\left[ -i\left( \Psi(f) - f' \frac{\ud \Psi}{\ud f} \right) \right] \,,\label{eq:leadingorderwf}
\ee
where here and in the rest of the paper, such derivatives with respect to the frequency will always be evaluated at $f$. Plugging this relation into~\eqref{eq:FDkernel}, we obtain
\begin{align}
	\tilde{s}(f) &\simeq \tilde{h}(f) \int \ud f' \, \exp\left[ i f' \frac{\ud \Psi}{\ud f} \right] \tilde{G}(f,f') \nn\\
	&= \tilde{h}(f) G\left( f, -\frac{1}{2\pi} \frac{\ud \Psi}{\ud f} \right) \,,\label{eq:leadingorderresponse}
\end{align}
which we can think of as a local evaluation of the kernel function $G(f,t)$ at a frequency-dependent effective time
\be\label{eq:deftf}
	\tf \equiv -\frac{1}{2\pi} \frac{\ud \Psi}{\ud f} \,.
\ee
It is worth noting that a shift in time of the time-domain signal will, by virtue of~\eqref{eq:shifttime}, be appropriately propagated to $t_{f}$. Because of the freedom of adding a linear term to $\Psi(f)$ by simply shifting the signal in time, no assumption can be made on the smallness of the first derivative of the phase, and this is really a leading order approximation.

>>>>

\usepackage{amsmath}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{latexsym}
\usepackage{graphicx}
\usepackage{bm}
\usepackage{color}
\usepackage{enumerate}
\usepackage{tabularx}
\usepackage{braket}

@@

\newcommand{\be}{\begin{equation}}
\newcommand{\ee}{\end{equation}}
\newcommand\ud{{\mathrm{d}}}
\newcommand\uD{{\mathrm{D}}}
\newcommand\calO{{\mathcal{O}}}
\newcommand\calM{{\mathcal{M}}}
\newcommand\calF{{\mathcal{F}}}
\newcommand\calT{{\mathcal{T}}}
\newcommand\calD{{\mathcal{D}}}
\newcommand\calA{{\mathcal{A}}}
\newcommand\bfx{\mathbf{x}}
\newcommand{\ov}[1]{\overline{#1}}
\newcommand{\ph}[1]{\phantom{#1}}
\newcommand{\cte}{\mathrm{cte}}
\newcommand{\nn}{\nonumber}
\newcommand{\hatk}{k}
\newcommand{\Hz}{\,\mathrm{Hz}}
\newcommand{\yr}{\,\mathrm{yr}}
\newcommand{\sinc}{\,\mathrm{sinc}}
\newcommand{\Msol}{M_{\odot}}
\newcommand{\Mchirp}{M_{c}}
\newcommand{\tf}{t_{f}}
\newcommand{\Tf}{T_{f}}
\newcommand{\tfd}{t_{f}^{d}}
\newcommand{\tfSPA}{t_{f}^{\rm SPA}}
\newcommand{\TfSPA}{T_{f}^{\rm SPA}}

@@

We present a Fourier-domain approach to modulations and delays of gravitational wave signals, a problem which arises in two different contexts. For space-based detectors like LISA, the orbital motion of the detector introduces a time-dependency in the response of the detector, consisting of both a modulation and a varying delay. In the context of signals from precessing spinning binary systems, a useful tool for building models of the waveform consists in representing the signal as a time-dependent rotation of a quasi-non-precessing waveform. In both cases, being able to compute transfer functions for these effects directly in the Fourier domain may enable performance gains for data analysis applications by using fast frequency-domain waveforms. Our results generalize previous approaches based on the stationary phase approximation for inspiral signals, extending them by including delays and computing corrections beyond the leading order, while being applicable to the broader class of inspiral-merger-ringdown signals. In the LISA case, we find that a leading-order treatment is accurate for high-mass and low-mass signals that are chirping fast enough, with errors consistently reduced by the corrections we derived. By contrast, low-mass binary black holes, if far away from merger and slowly-chirping, cannot be handled by this formalism and we develop another approach for these systems. In the case of precessing binaries, we explore the merger-ringdown range for a handful of cases, using a simple model for the post-merger precession. We find that deviations from leading order can give large fractional errors, while affecting mainly subdominant modes and giving rise to a limited unfaithfulness in the full waveform. Including higher-order corrections consistently reduces the unfaithfulness, and we further develop an alternative approach to accurately represent post-merger features.

@@

With the unprecedented recent gravitational-wave detections of coalescencing binary black holes and binary neutron stars, announced by the LIGO-Virgo collaboration~\cite{LIGO-theevent-2016, LIGO-christmasevent-2016, LIGO-O1BBH-2016, LIGO-BNSevent-2017}, gravitational-wave astronomy has entered its observational era. As LIGO prepares for even more sensitive observation runs, and with the recent expansion of the ground-based detectors network with Virgo~\cite{Virgo} (and eventually also KAGRA~\cite{KAGRA} and LIGO-India~\cite{INDIGO}), observations of such compact object coalescences are expected at an ever-increasing rate.

Moreover, the European Space Agency has recently selected the Laser Interferometer Space Antenna (LISA)~\cite{LISA17} to realize the ``Gravitational Universe'' science theme~\cite{elisa13} as the 3rd large space mission of its Cosmic Vision program, with a tentative launch around 2034. A technology demonstrator, LISA Pathfinder, has tested with great success some of the key mission technologies~\cite{LISAPathfinder2016, LISAPathfinder2018}. LISA will be able to detect and characterize, among several important gravitational wave source targets, comparable-mass binary black hole coalescences from cosmological distances over a wide range of masses. These will range from high-redshift observations of supermassive black hole binaries with $M\sim 10^{7} \Msol$ down to the observation of LIGO-type sources with $M\sim 10-100 \Msol$~\cite{Sesana16}.

@@

A complete representation of spin effects across parameter space in fast IMR templates still remains a frontier of gravitational wave signal modelling. In presence of misaligned spins, the system will endure precession of the orbital plane as it evolves, leading to modulations of the signal as seen by the observer~\cite{Apostolatos+94, Kidder95}. The presence of six degrees of freedom for the spins increases the dimensionality of the problem.

@@

Whereas LIGO and Virgo are typically sensitive to chirping binaries for a minute or less, LISA signals may accumulate over months or years. The response of a LISA-type instrument is thus time-dependent~\cite{Cutler97}. The motion and change of orientation of the detector constellation along its orbit lead to significant time variability in the form of a modulation and a varying delay. These effects then convey information about the localization of the gravitational-wave source in the sky. Direct time-domain implementation of the detector response is straightforward~\cite{Vallisneri04, Petiteau+08, CR02, RCP04}, but at a high computational cost for parameter-estimation analyses. To leverage the performance of state-of-the-art Fourier-domain IMR templates~\cite{BTB16,Khan+15}, we must efficiently process the signals through the time-dependent response of the dectector while staying in the Fourier domain.

@@

In the case of precessing binaries, applying the SPA directly to the modulated signal is prone to pathologies. In Refs.~\cite{KCY13,KCY14}, a formalism (called shifted uniform asymptotics or SUA) was introduced to go beyond the SPA and compute more accurately the modulation in the Fourier domain; however, this formalism still relies on the SPA for the underlying precessing-frame signal, and is as such limited to inspiraling signals. The simplified treatment of the precession response in the phenomenological waveforms of Ref.~\cite{Hannam+13} takes another approach, treating the precession modulation in the frequency domain by directly associating the Fourier frequency with the post-Newtonain orbital frequency. As we will explain, this corresponds to the zeroth-order approximation of the SUA.

In the case of the LISA response, the SPA provides a natural map from time-domain (for the orbit) to Fourier-domain (for the signal)~\cite{Cutler97}, and was used by many previous studies with inspiral waveforms. Ref.~\cite{Klein+15} included the orbital motion of the detector in the SUA treatment of precessing inspiral signals, within a low-frequency approximation of the constellation response. Consistently extending these previous approaches to the merger-ringdown part of the signals, while including the delays in the full LISA response at all frequencies, assessing and understanding the errors made along the way, is part of the objectives of this paper.

@@

The plan of the paper is as follows. In Sec.~\ref{sec:motivation}, we provide a general presentation of the problem of Fourier-domain modulation and introduce the relevant timescales for both the response of LISA-type detectors and the modulation of precessing signals. In Sec.~\ref{sec:formalism}, we present our general formalism, give its leading order approximation as well as higher-order corrections, introduce new timescales based on the Fourier-domain signal, and refine the previous results for both the quadratic-in-phase corrections and the treatment of the delays. We then apply our formalism to the response of the LISA detector in Sec.~\ref{sec:LISA}, and to the case of signals from precessing binaries in Sec~\ref{sec:precession}. We discuss and summarize our results in Sec.~\ref{sec:discussion}.

@@

@@

As is well known, GW signals decomposed in spin-weighted spherical harmonic components are smoothly varying functions of frequency in amplitude/phase form. An important consequence is that these signal components can be accurately represented in terms of a relatively coarsely sampled frequency grid. This property will also extend to the transfer functions, allowing one to keep a compact representation of the full signals. (See Appendix~\ref{app:notation} for details of our notation and conventions for Fourier transforms and spherical harmonics.)

@@

As a preliminary step, we recall here an approximation widely used for similar purposes in treating gravitational waves signals emitted by inspiraling binaries, the Stationary Phase Approximation (thereafter SPA, and also sometimes called the steepest-descent method). It applies in general to chirping signals, and we refer the reader to~\cite{FC93, CF94} for details. One first writes the time-domain signal in amplitude and phase form as $h(t) = a(t) e^{-2i\varphi(t)}$, where for gravitational wave signals $\varphi$ will correspond to the orbital phase, with the orbital frequency being $\omega = \dot{\varphi}$. In order to keep close to the notation used in the gravitational wave literature, we introduced a factor of 2 in the phase of the wave, which is appropriate for the dominant 22 harmonic of the signal. The approximation then applies to signals verifying the conditions\footnote{Note that the third condition is not always written explicitly, in particular not in Refs.~\cite{FC93, CF94}. It becomes important if we generalize $a$ to be an envelope function incorporating a modulation.}
\be\label{eq:conditionsSPA}
	\left| \frac{\dot{a}/a}{\omega} \right| \ll 1\,, \quad \left|\frac{\dot{\omega}}{\omega^{2}} \right| \ll 1\,, \quad \left| \frac{(\dot{a}/a)^{2}}{\dot{\omega}} \right| \ll 1 \,.
\ee
Since the integral~\eqref{eq:defFT} defining the Fourier transform is rapidly oscillatory unless the term $2\pi f t$ cancels the evolution of $-2\varphi(t)$, its support is well centered around the point of stationary phase. This determines the time-to-frequency correspondence in the stationary phase approximation, and leads to the definition of this time as an implicit function of frequency by the relation
\be\label{eq:deftfSPA}
	\omega(\tfSPA) = \pi  f \,.
\ee
For a chirping signal of increasing phase, $\omega>0$ and $\dot{\omega}>0$, there is a unique point of stationary phase, located in the positive frequency range $f>0$. Using the conditions above, one can formally expand the signal around $\tfSPA$ to quadratic order in time, according to
\begin{align}
	\tilde{h}_{\rm SPA} (f) &\simeq a(\tfSPA) \exp\left[2i\pi f \tfSPA-2i\varphi(\tfSPA) \right] \nn\\
	& \qquad \cdot \int \ud t \, e^{-i \dot{\omega} (\tfSPA) (t-\tfSPA)^{2}} \,.
\end{align}
Note that to be able to treat the amplitude as a constant in the integral above, we used the third condition in~\eqref{eq:conditionsSPA}. The resulting complex Gaussian integral yields\footnote{The expressions given here are valid for the dominant mode $h_{22}$ of the waveform. They can be generalized to other modes $h_{\ell m}$, $m\neq 0$ with phase $e^{-im\varphi}$ as follows: $t_{f}^{\rm SPA}$ is now such that $m\omega = 2\pi f$, $A$ acquires a factor $\sqrt{2/m}$, in $\Psi$ the term $2\varphi$ becomes $m\varphi$.}
\begin{subequations}
\begin{align}
	\tilde{h}_{\rm SPA}(f) &= A_{\rm SPA}(f) e^{-i\Psi_{\rm SPA}(f)} \,, \\
	A_{\rm SPA}(f) &= a(\tfSPA) \sqrt{\frac{\pi}{\dot{\omega}(\tfSPA)}} \,, \label{eq:ASPA} \\
	\Psi_{\rm SPA}(f) &= 2\varphi(\tfSPA) - 2\pi f \tfSPA + \frac{\pi}{4} \,. \label{eq:PsiSPA}
\end{align}
\end{subequations}
Ref.~\cite{Droz+99} evaluated the first correction to this approximation, within the context of post-Newtonian signals, and found that it can be considered as a term of the fifth post-Newtonian order, beyond the accuracy level of our current best models~\cite{BlanchetLiving}.

To understand the separation of timescales in the problem, it will be useful to have at hand the leading-order scaling laws for an inspiral (labeled the Newtonian order in the PN language). Although inaccurate for the purpose of waveform modelling, these leading-order estimates will give useful orders of magnitude of the relevant timescales in the inspiral. For a binary with masses $m_{1}, m_{2}$, we define the total mass $M=m_{1}+m_{2}$ and the symmetric mass ratio $\nu = m_{1}m_{2}/M^{2}$. Introducing a time of coalescence $t_{c}$, the relations between the orbital frequency and phase and the time to coalescence $t_{c} - t$ are then given by
\begin{subequations}\label{eq:omegaphiN}
\begin{align}
	\omega(t) &= \left[ \frac{256\nu}{5c^{5}} (GM)^{5/3} (t_{c}-t) \right]^{-3/8} \,, \\
	\varphi(t) &= -\left[ \frac{c^{3}}{5 G M \nu^{3/5}} (t_{c}-t) \right]^{5/8} \,.
\end{align}
\end{subequations}
As for the leading-order time-domain amplitude of the $22$ mode, we have~\cite{BlanchetLiving}
\be\label{eq:a22N}
	a_{22}^{\rm N} (t) = \frac{2 G M \nu v^{2}}{D c^{2}} \sqrt{\frac{16 \pi}{5}} \,,
\ee
where we set $v = (G M\omega/c^{3})^{1/3}$ and where $D$ is the luminosity distance to the observer. For this Newtonian inspiral, applying the SPA gives for the $22$ mode $\tilde{h}(f) = A_{\rm N}(f)e^{-i\Psi_{\rm N}(f)}$ with
\begin{subequations}\label{eq:SPAN}
\begin{align}
	A_{N}(f) &= \frac{G^{2}M^{2} \pi}{Dc^{5}} \sqrt{\frac{2\nu}{3}} v^{-7/2}\,, \label{eq:ASPAN}\\
	\Psi_{\rm SPA}^{\rm N}(f) &= \phi_{0} - 2\pi f t_{0} - \frac{3}{128\nu v^{5}} \,, \label{eq:PsiSPAN} 
\end{align}
\end{subequations}
where $v=(G M \pi f/c^{3})^{1/3}$ according to the SPA correspondence~\eqref{eq:deftfSPA}, and where $t_{0}, \phi_{0}$ are constants\footnote{For modes $h_{\ell m}$ with $m\neq 0$, the phase $\Psi$ acquires a factor $m/2$ and has to be evaluated at $(2/m)^{1/3} v$.}. It is also customary to rewrite the above relations in terms of the chirp mass $\Mchirp \equiv M\nu^{3/5}$, which is the only mass combination characterizing the signal at the leading PN order.

In this Newtonian, low-frequency limit one can check that each of the combinations \eqref{eq:conditionsSPA} indeed vanish at $\calO{(v^5)}$. but as the system approaches merger, these condition are no-longer satisfied. Near and after the merger, the SPA treatment is not applicable. Nonetheless, the SPA has been a useful workhorse in many gravitational analyses involving frequency domain transfer functions of the form~\eqref{eq:deftransfer}. The usual approach is simply to replace any time dependencies appearing the transfer function using \eqref{eq:deftfSPA}. We will reference the SPA treatment, as a familiar touchstone, as we develop a more general formalism.

@@

For the signal, then, we have
\be
	\tilde{h}(f-f') \simeq A(f) \exp\left[ -i\left( \Psi(f) - f' \frac{\ud \Psi}{\ud f} \right) \right] \,,\label{eq:leadingorderwf}
\ee
Plugging this relation into~\eqref{eq:FDkernel}, we obtain
\begin{align}
	\tilde{s}(f) &\simeq \tilde{h}(f) \int \ud f' \, \exp\left[ i f' \frac{\ud \Psi}{\ud f} \right] \tilde{G}(f,f') \nn\\
	&= \tilde{h}(f) G\left( f, -\frac{1}{2\pi} \frac{\ud \Psi}{\ud f} \right) \,,\label{eq:leadingorderresponse}
\end{align}
which we can think of as a local evaluation of the kernel function $G(f,t)$ at a frequency-dependent effective time
\be\label{eq:deftf}
	\tf \equiv -\frac{1}{2\pi} \frac{\ud \Psi}{\ud f} \,.
\ee
It is worth noting that a shift in time of the time-domain signal will, by virtue of~\eqref{eq:shifttime}, be appropriately propagated to $t_{f}$. Because of the freedom of adding a linear term to $\Psi(f)$ by simply shifting the signal in time, no assumption can be made on the smallness of the first derivative of the phase, and this is really a leading order approximation.
